Towards a free/open-source universal-dependency treebank for Kazakh

### 2015-06-21 2015-turklang-kazdep/paper.tex
figs
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60933 72bbbca6-d526-0410-a7d9-f06f51895060

We have uniformly annotated the copula as a leaf node with the predicate, or adverbial
as the head of the structure. For certain structures this is convenient, such as the
bare copula in phrases like \ref{ex:cop1} or \ref{ex:cop2}, but for phrases where
the copula is part of an embedded clause this is not necessarily the most  
effective choice. In \ref{ex:cop3}, the copula holds all the morphological information,
agreement with the subject and accusative marking for the embedded clause.

@@

        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.3cm]
                                Айгүл \& студент \& · \\
                                \gmk{np} \& \gmk{n} \& \gmk{cop} \\
                                \tgloss{Aygül} \& \tgloss{student} \& \tgloss{is} \\
                        \end{deptext}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{xxx }\label{fig:compound}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.3cm]
                                Айгүл \& студент \& болған \\
                                \gmk{np} \& \gmk{n} \& \gmk{past} \\
                                \tgloss{Aygül} \& \tgloss{student} \& \tgloss{was}\\
                        \end{deptext}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{xxx.}\label{fig:cop2}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& Азаматтың \& қайда \& екенін \& білмейді. \\
                                \gmk{np} \& \gmk{np.gen} \& \gmk{adv} \& \gmk{cop} \& \gmk{neg.aor} \\
                                \tgloss{Aygül} \& \tgloss{Azamat} \& \tgloss{where} \& \tgloss{being} \& \tgloss{knows not} \\
                        \end{deptext}
                        \depedge{5}{1}{\udlabel{subj}}
                        \depedge{3}{2}{\udlabel{subj}}
                        \depedge{5}{3}{\udlabel{ccomp}}
                        \depedge{3}{4}{\udlabel{cop}}
                \end{dependency}
                \caption{xxx.}\label{fig:cop3}
        \end{subfigure}

        \caption{Dependency trees of copula constructions.}
\end{figure}

@@

	\begin{dependency}[theme = simple, font = \small]
	   \begin{deptext}[column sep=0.08cm]
%               1         2         3          4              5                6              7                8           9             10
             Барлық \& адамдар \& тумысынан \& азат      \& және         \& қадір-қасиеті \& мен          \& құқықтары  \& тең       \& болып     \& дүниеге \& келеді \\
                    \&         \&           \& \gmk{adj} \& \gmk{cnjcoo} \& \gmk{n}       \& \gmk{cnjcoo} \& \gmk{n.pl} \& \gmk{adj} \& \gmk{cop} \& \& \\
                    \&         \&           \& \tgloss{free} \& \tgloss{and}  \& \tgloss{dignity}  \&  \tgloss{and} \& \tgloss{rights} \&  \tgloss{equal} \& \tgloss{being} \& \& \\
		\end{deptext}
		\depedge{9}{4}{\udlabel{conj}}
		\depedge{9}{5}{\udlabel{cc}}
		\depedge{8}{6}{\udlabel{conj}}
		\depedge{8}{7}{\udlabel{cc}}
		\depedge{9}{8}{\udlabel{subj}}
		\depedge{10}{9}{\udlabel{cop}}
		\depedge{12}{10}{\udlabel{advcl}}
	\end{dependency}
	\caption{Coordination}\label{fig:coord}

>>>>

We have uniformly annotated the copula as a leaf node with the predicate, or adverbial
as the head of the structure. For certain structures this is convenient, such as the
bare copula in phrases like \ref{fig:cop1} or \ref{fig:cop2}, but for phrases where
the copula is part of an embedded clause this is not necessarily the most  
effective choice. In \ref{fig:cop3}, the copula holds all the morphological information,
agreement with the subject and accusative marking for the embedded clause.

@@

        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& студент \& · \\
                                \gmk{np} \& \gmk{n} \& \gmk{aor} \\
                                \gloss{Aygül} \& \gloss{student} \& \gloss{is} \\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{2}{\udlabel{root}}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{Non-surfaced copula.}\label{fig:cop1}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& студент \& болған \\
                                \gmk{np} \& \gmk{n} \& \gmk{past} \\
                                \gloss{Aygül} \& \gloss{student} \& \gloss{was}\\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{2}{\udlabel{root}}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{Copula in past tense.}\label{fig:cop2}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& оның \& қайда \& екенін \& білмейді. \\
                                \gmk{np} \& \gmk{gen} \& \gmk{adv} \& \gmk{cop} \& \gmk{neg.aor} \\
                                \gloss{Aygül} \& \gloss{he} \& \gloss{where} \& \gloss{being} \& \gloss{knows not} \\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{5}{\udlabel{root}}
                        \depedge{3}{2}{\udlabel{subj}}
                        \depedge{5}{3}{\udlabel{ccomp}}
                        \depedge{3}{4}{\udlabel{cop}}
                \end{dependency}
                \caption{Copula in an embedded clause.}\label{fig:cop3}
        \end{subfigure}
        \caption{Dependency trees of copula constructions.}
\end{figure}

@@

        \begin{dependency}[theme = simple, font = \small]
           \begin{deptext}[column sep=0.08cm]
%               1         2         3          4              5          6          7        8
                Олар \& Финляндия, \& Швеция \& және \& Эстонияидан \& кейін \& Ресейге \& барды. \\
                \gmk{prn} \& \gmk{nom} \& \gmk{nom} \& \gmk{cc} \& \gmk{abl} \& \gmk{post} \& \gmk{dat} \& \gmk{past} \\
                \gloss{They} \& \gloss{Finland}, \& \gloss{Sweden} \& \gloss{and} \& \gloss{Estonia} \& \gloss{after} \& \gloss{Russia} \& \gloss{visited.} \\
                \end{deptext}
                \depedge{5}{3}{\udlabel{conj}}
                \depedge{5}{4}{\udlabel{cc}}
                \depedge{5}{2}{\udlabel{conj}}
                \depedge{5}{6}{\udlabel{case}}
                \depedge{8}{5}{\udlabel{nmod}}
                \depedge{8}{7}{\udlabel{nmod}}
        \end{dependency}
        \caption{Coordination: All conjucts are attached to the final conjunct, which is the head of the coordinated phrase.}\label{fig:coord}
